Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-05 doc/paper/paper.tex
MBE Submission

\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}
\usepackage{cleveref}
\usepackage{array}

\newcommand{\defn}[1]{\textit{#1}}
\newcommand{\software}[1]{\textit{#1}}

\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
\newcommand{\sett}[1]{\{#1\}}
\renewcommand{\emptyset}{\varnothing}
\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\star}[1]{#1^*}
\renewcommand{\d}{\mathrm{d}\,}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\DeclareMathOperator{\Exponential}{Exponential}
\DeclareMathOperator{\Uniform}{Uniform}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}

\DeclareMathOperator{\inc}{in}
\DeclareMathOperator{\out}{out}

\newcommand{\N}{\mathcal{N}}
\renewcommand{\L}{\mathcal{L}}

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}
\newacronym{HPD}{HPD}{highest posterior density}
\newacronym{GTR}{GTR}{general time-reversible}
\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
\newacronym{IDU}{IDU}{injection drug user}

\graphicspath{{../figures/}}
\newcommand{\tablepath}{../tables}
\def\figwidth{0.8\textwidth}

\title{Phylodynamic inference of contact network parameters through approximate
Bayesian computation}
\author{Rosemary M. McCloskey \and Richard H. Liang \and Art F.Y. Poon}

\addbibresource{papers.bib}

\frenchspacing
\onehalfspacing
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\section*{Background}

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, but for sexually transmitted diseases, either sexual or
blood-to-blood contact is required. The set of links between individuals along
which transmission can occur is called the contact
network~\autocite{klovdahl1985social, morris1993epidemiology}. The structure of
the contact network underlying an epidemic can profoundly impact the speed and
pattern of the epidemic's expansion. Network structure can influence the
prevalence trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the intra-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2011estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, although
sophisticated methods have been developed for fitting complex population
genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics}, suggesting that such statistical
inference might be possible~\autocite{welch2011statistical}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[although there has been
some disagreement, see][]{jones2003assessment}. Under these models, nodes with
a high number of contacts attract new connections at an elevated rate. Networks
produced by preferential attachment have a power-law degree distribution,
meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
for some constant $\gamma$. These networks are also referred to as
``scale-free''. The first contact network model incorporating preferential
attachment was introduced by \textcite{barabasi1999emergence}, and is now
referred to as the \gls{BA} model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. These new edges are joined to
existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
that nodes of high degree tend to attract more connections (in the original
paper, only $\alpha = 1$ was investigated).

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} analysed the degree distribution
of an approximate transmission network, estimated based on genetic similarity
and estimated times of infection, relating 60\% of HIV infected \gls{MSM} in
the United Kingdom. The transmission network is a subgraph of the contact
network which includes only those edges which have already led to a new
infection. The authors found that a Waring distribution, which is produced by a
different preferential attachment model, was a good fit to their estimated
network.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using \gls{MCMC}. Rather than calculating explicit likeilihoods, \gls{MCMC}
uses likelihood \emph{ratios} in which the normalizing constant cancel out.
Unfortunately, it is generally difficult to explicitly calculate the likelihood
of an observed transmission tree under a contact network model, even up to a
normalizing constant. To do so, it would be necessary to integrate over all
possible networks, and also over all possible labellings of the internal nodes
of the transmission tree. While it is not known (to us) whether such
integration is tractable, a simpler alternative is offered by likelihood-free
methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
calculating the likelihood may be impossible, generating simulated datasets
according to a model is often straightforward. If our model fits the data well,
the simulated data it produces should be similar to the observed data. More
formally, if $D$ is the observed data, the posterior distribution $f(\theta
\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
specific case when $\rho$ is a kernel function, the approach is known as
kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree, using the tree
kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
of the \gls{BA} model on a variety of simulated and real data sets. Our results
show that the attachment power parameter $\alpha$ can be inferred with
reasonable accuracy, and can vary considerably between epidemics from different
settings.

\section*{Methods}

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
in the same fashion as several previous studies
\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
that our implementation was correct, we reproduced Figure 1A of
\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
unbalancedness of transmission trees simulated over four network models at
various levels of pathogen transmissibility. Our program is freely available at
\url{https://github.com/rmcclosk/netabc}.

We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
denotes the total number of nodes in the network, and $I$, which is the number
of infected nodes at which to stop the simulation and sample the transmission
tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
they became infected at a rate proportional to their numbers of infected
neighbours, and never recovered. For all analyses, the transmission trees'
branch lengths were scaled by dividing by their mean. We used the 
\software{igraph} library's implementation of the \gls{BA}
model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster.

\subsection*{Kernel classifiers}

We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
whether the parameters of the \gls{BA} model had an effect on tree shape. We
simulated 100 networks under each of three different values of $\alpha$: 0.5,
1.0, and 1.5 (300 networks total). The other parameters were fixed to the
following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
500 tips was simulated over each network (300 transmission trees total). The
300 trees were compared pairwise with the tree kernel to form a $300 \times
300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
factor''), and $\sigma$ (the ``radial basis function
variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4
respectively. We constructed a \gls{kSVM} classifier for $\alpha$ using the
\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
accuracy with 1000 two-fold cross-validations.

Three similar experiments were performed for the other \gls{BA} model
parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
$m$, and $N$.

We repeated these four cross-validations with different values of $\lambda$
(0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well
as on trees with differing numbers of tips (100, 500, and 1000) and in
epidemics of differing size $I$ (500, 1000, and 2000). When evaluating the
classifier for $I$, we did not consider trees with 1000 tips, because one of
the tested $I$ values was 500, and the number of tips cannot be larger than
$I$.

For each of the four parameters, we also tested univariate classifier based on
Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the
\gls{nLTT} statistic~\autocite{janzen2015approximate}.

\subsection*{ABC simulations}

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. To check that our implementation was correct, we
applied it to the same mixture of Gaussians used by
\citeauthor{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see \cref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps.

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1,
1.5}. This produced a total of 24 parameter combinations $\times$ three trees
per combination = 72 trees total. The adaptive \gls{ABC} algorithm was applied
to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
Uniform(0, 2), and $(N, I)$ jointly uniform on the triangular region \{$500
\leq N \leq 15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
were Gaussian, with variance equal to twice the empirical variance of the
particles. The algorithm was run with 1000 particles, 5 simulated datasets per
particle, and the ``quality'' parameter controlling the decay rate of the
tolerance $\varepsilon$ set to 0.95. We used the same stopping criterion as
\citeauthor{del2012adaptive}, namely when the \gls{MCMC} acceptance rate
dropped below 1.5\%. Point estimates for the parameters were obtained by taking
the highest point of an estimated kernel density on the final set of particles,
using the \software{density} function with the default parameters in
\software{R}. \gls{HPD} intervals were calculated with the
\software{HPDinterval} function from the \software{R} package
\software{coda}~\autocite{plummer2006coda}.

>>>>

\documentclass[nogrid]{MBE}
\usepackage{url}

% Mol Biol Evol stuff
\jshort{mst}
\volname{}
\jvolume{0}
\jvol{}
\jissue{0}
\pubyear{2016}
\mstype{Article}
\artid{???}
\access{???}
\begin{document}

\title[Contact networks and ABC]{Phylodynamic inference of contact network
parameters through approximate Bayesian computation}
\author[McCloskey, Liang, and Poon]{Rosemary M. McCloskey$^{*,1}$, 
    Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$}
\address{$^1$BC Centre for Excellence in HIV/AIDS, 608-1081 Burrard Street, Vancouver, Canada \\
$^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
\history{Submitted 5 April 2016}
\coresp{E-mail: rmccloskey@cfenet.ubc.ca.}
\editor{}
\abstract{\input{abstract}}
\keyword{Phylogenetics, phylodynamics, approximate Bayesian computation,
contact network, transmission tree, human immunodeficiency virus.}

@@

\section{Introduction}

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, while for sexually transmitted diseases, \newpage\noindent %hack
sexual or in some cases blood-to-blood contact is required. The population
together with the set of links between individuals along which transmission can
occur is called the contact network~\citep{klovdahl1985social,
morris1993epidemiology}. The structure of the contact network underlying an
epidemic can profoundly impact the speed and pattern of the epidemic's
expansion. Network structure can influence the prevalence
trajectory~\citep{o2011contact} and epidemic
threshold~\citep{barthelemy2005dynamical}, in turn affecting the estimates of
quantities such as effective population size~\citep{goodreau2006assessing}.
From a public health perspective, contact networks have been explored as tools
for curtailing epidemic spread, by way of interventions targeted to
well-connected nodes~\citep{wang2015targeting}. True contact networks are a
challenging type of data to collect, requiring extensive epidemiological
investigation~\citep{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the inter-host viral
phylogeny~\citep{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\citep{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\citep{stadler2012estimating}, to the degree of
clustering~\citep{hughes2009molecular}, to the elevated transmission risk
during acute infection~\citep{volz2012simple}. On the other hand, although
sophisticated methods have been developed for fitting complex population
genetic models to phylogenies~\citep{rasmussen2014phylodynamic}, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\citep{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
suggesting that such statistical inference might be
possible~\citep{welch2011statistical}.

Survey-based studies of sexual networks~\citep{liljeros2001web,
schneeberger2004scale, colgate1989risk} have found that these networks tend to
have a degree distribution which follows a power law \citep[although there has
been some disagreement, see][]{handcock2004likelihood}. That is, the number of
nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
$\gamma$. These networks are also referred to as ``scale-free''. One process by
which scale-free networks can be generated is preferential attachment, where
nodes with a high number of contacts attract new connections at an elevated
rate. The first contact network model incorporating preferential attachment was
introduced by \citet{barabasi1999emergence}, and is now referred to as the
Barab\'asi-Albert (BA) model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. In the most commonly studied
formulation, these new edges are joined to existing nodes of degree $k$ with
probability proportional to $k$, so that nodes of high degree tend to attract
more connections. \citeauthor{barabasi1999emergence} suggested an extension
where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
some non-negative constant $\alpha$, and we use this extension here.

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \citet{britton2002bayesian} develop a Bayesian
approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\citep{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\citet{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \citet{volz2007susceptible, volz2008sir} developed differential
equations describing the spread of a susceptible-infected (SI) epidemic on
static and dynamic contact networks with several degree distributions, which
could in principle be used for inference if observed incidence trajectories
were available. \citet{brown2011transmission} analysed the degree distribution
of an approximate transmission network, estimated based on genetic similarity
and estimated times of infection, relating 60\% of HIV-infected men who have
sex with men (MSM) in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection. The authors found that a Waring distribution,
which is produced by a more sophisticated preferential attachment model, was a
good fit to their estimated network.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
which the normalizing constants cancel out. Unfortunately, it is generally
difficult to explicitly calculate the likelihood of an observed transmission
tree under a contact network model, even up to a normalizing constant. To do
so, it would be necessary to integrate over all possible networks, and also
over all possible labellings of the internal nodes of the transmission tree.
While it is not known (to us) whether such integration is tractable, a simpler
alternative is offered by likelihood-free methods, namely approximate Bayesian
computation (ABC). ABC leverages the fact that, although calculating the
likelihood may be impractical, generating simulated datasets according to a
model is often straightforward. If our model fits the data well, the simulated
data it produces should be similar to the observed data. More formally, if $D$
is the observed data, the posterior distribution $f(\theta \mid D)$ on model
parameters $\theta$ is replaced as the target of statistical inference by
$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\citep{sunnaaker2013approximate}. In the
specific case when $\rho$ is a kernel function, the approach is known as
kernel-ABC~\citep{nakagome2013kernel, poon2015phylodynamic}.

Here, we develop a method using kernel-ABC to estimate the parameters of
contact network models from observed phylogenetic data. The distance function
we use is the tree kernel developed by \citet{poon2013mapping}, which computes
a weighted dot product of the trees' representations in the space of all
possible subset trees. We apply the method to investigate the parameters of the
BA network model on a variety of simulated and real datasets. Our results show
that some network parameters can be inferred with reasonable accuracy, while
others have a minimal detectable impact on tree shape and therefore cannot be
estimated accurately. We also find that these parameters can vary considerably
between real epidemics from different settings.

\section{New Approaches}

We have developed a kernel-ABC-based method to perform statistical inference of
contact network parameters from a transmission tree estimated from an observed
viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
algorithm for ABC developed by \citet{del2012adaptive}. The SMC algorithm keeps
track of a population of parameter ``particles'', which are initially sampled
from the parameters' joint prior distribution. Several datasets are simulated
under the model of interest for each of the particles. In this case, the
datasets are transmission trees, which are generated by a two-step process.
First, a contact network is simulated according to the network model being fit.
Second, a transmission tree is simulated over that network with a Gillespie
simulation algorithm~\citep{gillespie1976general}, in the same fashion as
several previous studies \citep[\textit{e.g.}][]{robinson2013dynamics,
leventhal2012inferring}. The particles are weighted according to the similarity
between their associated simulated trees and the observed tree. To quantify
this similarity, we used the tree kernel developed by \citet{poon2013mapping}.
Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
and, if the move is accepted, simulating new datasets under the new parameters.
When a particle's weight drops to zero, because its simulated trees are too
dissimilar to the observed tree, the particle is dropped from the population,
and eventually replaced by a resampled particle with a higher weight. As the
algorithm progresses, the population converges to a Monte Carlo approximation
of the ABC target distribution, which is assumed to approximate the desired
posterior~\citep{del2012adaptive, sunnaaker2013approximate}.

A computer program implementing our method is freely available at
\url{https://github.com/rmcclosk/netabc} (last accessed April 3, 2016).

\section{Results}

\subsection{Kernel classifiers}

We investigated the parameters of the BA network
model~\citep{barabasi1999emergence}. In addition to $m$ and $\alpha$ (see
Introduction), we considered $N$, which denotes the total number of nodes in
the network, and $I$, which is the number of infected nodes at which to stop
the simulation and sample the transmission tree. To examine the effect of these
parameters on tree shape, we simulated transmission trees under different
parameter values, calculated pairwise tree kernel scores between them, and
attempted to classify the trees using a kernel support vector machine (kSVM).
We also tested classifiers based on Sackin's index~\citep{shao1990tree} and the
normalized lineages-through-time (nLTT)
statistic~\citep{janzen2015approximate}. Accuracy of the kSVMs varied based on
the parameter being tested (Figure~\ref{fig:rsquared}, left). Classifiers based
on two other tree statistics, the nLTT and Sackin's index, generally exhibited
worse performance than the tree kernel, although the magnitude of the disparity
varied between the parameters (Figure~\ref{fig:rsquared}, centre and right).
The results were largely robust to variations in the tree kernel
meta-parameters $\lambda$ and $\sigma$ (Figures~S3-S6).

When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
    0.92,
compared to 
    0.56
for the nLTT-based SVM, and
    0.75
for the linear regression against Sackin's index. There was little variation
about the mean for different tree and epidemic sizes. No classifier could
accurately identify the $m$ parameter in any epidemic scenario, with average
$R^2$ values of 
  0.12 for kSVM,
  0.01 for the nLTT, and
  0.06
for Sackin's index. Again, there was little variation in accuracy between
epidemic scenarios, although the accuracy of the kSVM was slightly higher
on 1000-tip trees.

The accuracy of classifiers for $I$ varied significantly with the number of
tips in the tree. For 100-tip trees, the average $R^2$ values were
  0.7,
  0.55, and
  0.02
for the tree kernel, nLTT, and Sackin's index respectively. For 500-tip
trees, the values increased to
  0.93,
  0.83, and
  0.07.
Finally, the performance of classifiers for $N$ depended heavily on the
epidemic scenario. The $R^2$ of the kSVM classifier ranged from
  0.08
for the smallest epidemic and smallest sample size, to
  0.82
for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
  0.01
to
  0.54.
Sackin's index did not accurately classify $N$ in any scenario, with an average
$R^2$ of
  0.03
and little variation between scenarios.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=\textwidth]{kernel-rsquared.eps}
  \vspace{6pt}
  \caption{
      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
      using nLTT (centre), and linear regression using Sackin's index
      (right) for BA model parameters. Kernel meta-parameters were set to
      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
      simulated transmission trees over networks with three different values of
      the parameter being tested. Vertical lines are empirical 95\% confidence
      intervals based on 1000 two-fold cross-validations.
  }
  \label{fig:rsquared}
\end{figure*}

\subsection{ABC simulations}

Figure~\ref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
estimates of the BA model parameters obtained with kernel-ABC on simulated
data. The estimates shown correspond only to the simulations where the $m$
parameter was set to 2, however the results for $m = 3$ and $m = 4$ were
similar (Figures S7 and S8). Average boundaries of 95\% HPD intervals are given
in Table~\ref{tab:abchpd}.

The accuracy of the parameter estimates obtained with kernel-ABC
paralleled the results from the kSVM classifier. Of the four parameters,
$\alpha$ was the most accurately estimated, with point estimates having a
median [IQR] absolute error of 
    0.11 
    [0.05 - 
    0.18].
The errors when the true value of $\alpha$ was zero were significantly greater
than those for the other values 
    (Wilcoxon rank-sum test, $p$ = $6\!\times\!10^{-4}$),
but did not vary across the true values of the other parameters (one-way
ANOVA). Estimates for $I$ were also relatively accurate, with point estimate
errors of
    306 
    [108 - 
    607].
These errors were significantly higher when the true value of $\alpha$ was
at least 1
    (Wilcoxon rank-sum test, $p$ = $6\!\times\!10^{-4}$)
and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
did not affect the estimates of $I$ (one-way ANOVA).

The $m$ parameter was estimated correctly in only
    37 \%
of simulations. Oddly, the error in the estimated $m$ was higher for integer
values of $\alpha$ (\text{i.e.}~0 and 1) than non-integer values 
    (Wilcoxon rank-sum test, $p$ = 0.007).
The true values of the other parameters did not significantly affect the
estimates of $m$ (both one-way ANOVA). Finally, the total number of nodes $N$
was consistently over-estimated by about a factor of two
    (error 6588 
    [4214 - 
     8284]).
No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).

\begin{figure*}[ht]
  \centering
  \includegraphics[width=\textwidth]{abc-point-estimate-m2.eps}
  \vspace{6pt}
  \caption{
    Point estimates of BA model parameters obtained by running kernel-ABC
    on simulated phylogenies without training, for simulations with $m = 2$.
    Dotted lines indicate true values, and limits of the $y$-axes are regions
    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
    true values in simulations. (B) Estimates for $m$ and $N$, which were held
    fixed in these simulations, against true values of $\alpha$.
  }
  \label{fig:abcpt}
\end{figure*}

\begin{table*}[ht]
  \centering
    \begin{tabular}{lr>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}}
      \hline
    Parameter & True value & Mean point estimate & Mean HPD lower bound & Mean HPD upper bound \\ 
      \hline
    $\alpha$ & 0.0 & 0.24 & 0.02 & 0.73 \\ 
       & 0.5 & 0.42 & 0.02 & 0.81 \\ 
       & 1.0 & 0.97 & 0.61 & 1.11 \\ 
       & 1.5 & 1.48 & 1.26 & 1.83 \\ 
      $I$ & 1000 & 1155.68 & 598.68 & 2402.84 \\ 
       & 2000 & 2646.07 & 1182.31 & 4058.13 \\ 
      $m$ & 2 & 2.92 & 1.75 & 4.92 \\ 
       & 3 & 3.33 & 1.96 & 4.92 \\ 
       & 4 & 3.62 & 1.88 & 5.00 \\ 
      $N$ & 5000 & 10962.61 & 2732.55 & 14701.87 \\ 
       \hline
    \end{tabular}
  \caption{
      Average maximum \textit{a posteriori} point estimates and 95\% highest
      posterior density (HPD) interval widths for BA model parameter estimates
      obtained with kernel-ABC. Three transmission trees were simulated under
      each combination of the listed parameter values, and the parameters were
      estimated with kernel-ABC without training.
  }
  \label{tab:abchpd}
\end{table*}

The dispersion of the ABC approximation to the posterior also varied between
the parameters, with narrower HPD intervals for the parameters with more
accurate point estimates (Table~\ref{tab:abchpd}). Figure~\ref{fig:abcex} shows
the distributions for for one simulation (equivalent plots for all the
simulations can be found in Supplementary Figure S10). HPD intervals around $\alpha$
and $I$ were narrow relative to the region of nonzero prior density, whereas
the intervals for $m$ and $N$ were more widely dispersed.

\begin{figure*}[ht]
    \centering
  \includegraphics[width=\textwidth]{abc-posterior-example.eps}
  \vspace{6pt}
  \caption{
    Marginal posterior distributions of BA model parameters estimated
    with kernel-ABC for a single simulated transmission tree. Dotted
    lines and shaded polygon indicate true values.
  }
  \label{fig:abcex}
\end{figure*}

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1 
  [0.07-
   1.07];
$I$,
  1170 
  [506 -
   1616];
$m$,
  5 
  [2 -
   5];
$N$,
  \ensuremath{1.033\times 10^{4}} 
  [2408-
   \ensuremath{1.4974\times 10^{4}}].
To test the effect of sampling bias, we sampled one transmission tree in a
peer-driven fashion, where the probability to sample a node was twice as high
if one of its peers had already been sampled. The parameters for this
experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
estimated values were
$\alpha$, 
  0.48 
  [0.03 -
   0.82];
$I$,
  2468 
  [1281 -
   3938];
$m$,
  3 
  [2 -
   5];
$N$,
  \ensuremath{1.0738\times 10^{4}} 
  [3106 -
   \ensuremath{1.4783\times 10^{4}}].
Both of these results were in line with estimates obtained on other simulated
datasets (Table~\ref{tab:abchpd}).

\subsection{Real data}
